A Theory for Database Transactions with no Database

### 2019-03-18 2018 Transactions/articleTransactions.tex
I have made things consistent with "the state of my mind".

\section{Growth}
\label{sct:Growth}
	A DIS grows by adding events to it.
	We describe each time an event $e$ happens as a mapping, $\id{happens}_{P,e}$, from one history structure to another:
\begin{equation}
\begin{array}{l}
	P\subseteq\Events\ \wedge\ e\notin E\\
	\ \ \Rightarrow\ \ \id{happens}_{P,e}(\la \Events, \id{prec}\ra)\ =\ \la \Events\cup\{e\}, \id{prec}\cup\{(e_1,e) | e_1\in P\}\ra
\end{array}
\label{req:def happens}
\end{equation}
	This happening adds one event to $\Events$ and enlarges relation $\id{prec}$ with one pair for every immediate predecessor of $e$.
	The immediate predecessors of $e$ are specified in set $P$, which we call the predecessor set.
	We need the predecessor set to cater for the distributed nature of a DIS.
\begin{lemma}
\label{lemma:happens preserves history structure}
	If $\la \Events, \id{prec}\ra$ is a history structure and $P\subseteq\Events$ and $e\notin\Events$
	then $\id{happens}_{P,e}(\la \Events, \id{prec}\ra)$ is a history structure.
\end{lemma}
	Lemma~\ref{lemma:happens preserves history structure} is a preservation property.
	It says that $\id{happens}_{P,e}$ preserves the history structure,
	provided that $P\subseteq\Events$ and $e\notin\Events$.
	This lemma has been proven in proof assistant Isabelle/HOL.
	To understand why $P\subseteq\Events$ is necessary,
	let $\la \Events', \id{prec}'\ra=\id{happens}_{P,e}(h)$.
	Now suppose event $e_1\in P$ but $e_1\notin\Events$.
	Equation~\ref{req:def happens} tells us that $(e_1,e)\in\id{prec}'$.
	However, this violates requirement~\ref{req:precInE}.
	So it is necessary that every $e_1\in P$ is an element of $\Events$.
	Similarly, $e\notin\Events$ is a necessary condition,
	because any $e\in\Events$ violates the antisymmetry of $\id{prec}^+$ (requirement~\ref{req:precPlus asy}).

	We extend the notion of subset to historic structures and use the subset symbol $\subseteq$ for that purpose.
	Let us represent the fact that $h$ is a {\em historic substructure} of $h'$ by:
\begin{equation}
\label{def:historic substructure}
	\la \Events, \id{prec}\ra\ \subseteq\ \la \Events', \id{prec}'\ra\ \Leftrightarrow\ \Events\subseteq\Events'\ \wedge\ \id{prec}\subseteq\id{prec}'
\end{equation}
	From lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{happens}_{P,e}(h)=h'$ implies that $h\subseteq h'$.

@@

	Let us reconsider the relation $\id{prec}$ for a moment and define $\preceq$ as follows:
	Requirement~\ref{req:precPlus asy} from section~\ref{sct:History} ensures that $\preceq$ is reflexive, antisymmetric, and transitive.
	Let us introduce operator $\curlyvee$ (the {\em meet} of two events)
\begin{equation}
	e\ =\ e_1\curlyvee e_2\ \ \Leftrightarrow\ \ e\ \preceq\ e_1\ \wedge\ e\ \preceq\ e_2
\end{equation}
	If $\la \Events, \id{prec}, \id{state}\ra$ satisfies definition~\ref{req:recursive def occurs},
	we can prove
\begin{equation}
		e_1,e_2\in\Events\ \Rightarrow\ (e_1\curlyvee e_2)\in\Events
\end{equation}
	This means that $\la\Events,\preceq\ra$ is an order-theoretic meet-semilattice.
	From order theory, we know that $\curlyvee$ is idempotent, commutative, and associative.

	Having defined history structures and the way they grow, let us proceed to determine requirements for states.

\section{States}
	This section defines the requirements for predictable behavior of a distributed information system (DIS).
	We consider behavior to be predictable if every state can be computed unambiguously from its history.
	For that purpose we introduce the notion of state.

	A history structure can be perceived as the (global) state of an information system at any point in time,
	even though that global state is observable only in its local parts.

	Let $\mathbb S$ represent the set of all conceivable states.
	A {\em stateful history structure} is a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
	$\la \Events, \id{prec}\ra$ is a history structure and $\id{state}:\Events\rightarrow{\mathbb S}$ is a function
	that maps every event in $\Events$ to a state.
	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful history structure.
	Let $s=\id{state}(e)$ represent the fact that $s$ is the state immediately after event $e$ has happened.
	Or shorter: $s$ is the state after event $e$.
	Instead of $s=\id{state}(e)$, we may also write $(s,e)\in\id{state}$ because a function is a univalent and total relation
	(and therefore a set of pairs).
	A {\em current state} is the state after a current event
	and a {\em past state} is the state after a past event.

	In a stateful history structure, the occurrence of an event $e$ defines a new history structure.
	This is the way systems evolve over time.
	So we define the occurrence of an event as a mapping, $\id{occurs}_{P,e,s}$, from one SHS to another SHS:
\begin{equation}
\begin{array}{l}
	\id{occurs}_{P,e,s}(\la \Events, \id{prec}, \id{state}\ra)\\
	\ =\ \la \Events\cup\{e\},\ \id{prec}\cup\{(e_1,e) | e_1\in P\},\ \id{state}\cup\{(e,s)\}\ra
\end{array}	
\label{eqn:def occurs}
\end{equation}
	This definition extends the definition of $\id{happens}$ (eqn.~\ref{req:def happens}).
	By using lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
	the stateful history structure.
	Also, we will use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em stateful historic substructure} of $h'$.
	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.
	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful history structure.
	To describe the growth of a stateful history structure by the event,
	we define $\id{SHS}$ recursively using $\id{occurs}_{P,e,s}$ as recurring step:
\begin{eqnarray}
	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
	\ \ \Rightarrow\ \id{SHS}(\id{occurs}_{P,e,s}\la \Events, \id{prec}, \id{state}\ra)
\end{array}
\label{req:recursive def occurs}
\end{eqnarray}

\section{Transactions}
\label{sct:Transactions}
	In practice, a new state is computed from the current state by specifying a change, $\Delta$.
	Such changes are called {\em transactions}.
	To describe a DIS in terms of changes, we need operators to compute a new state ($s_2$) from a given state ($s_1$).
	In fact, we need two operators, $\oplus$ and $\ominus$, such that
\begin{equation}
	s_2\ =\ s_1\oplus(s_2\ominus s_1)
\label{req:oplus and ominus}
\end{equation}
	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
	This difference represents a change.
	We use $s_1\oplus\Delta$ to compute a new state from an existing state $s_1$ and a change $\Delta$.
	In a stateful history structure we can represent transactions by means of a function $\delta$,
	where $\delta(e_1,e_2)$ represents the change of state from event $e_1$ to $e_2$.
\begin{equation}
	\delta(e_1,e_2)\ =\ \id{state}(e_2)\ominus\id{state}(e_1)
\label{def:delta}
\end{equation}
	A transaction caused by event $e$ can be described by:
&&	h'=\id{occurs}_{P,e,s}(h)\ \Rightarrow\ \begin{array}[t]{@{}ll}
		\id{state}(e)\ =\ s&\wedge\\
		\forall e_1\in P: s\ =\ \id{state}(e_1)\oplus\delta(e_1,e)
\label{req:transaction}

@@

	\ \ =\ \begin{array}[t]{@{}l@{}c@{}ll}

@@

		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{state}\cup\{(e,s_1\oplus\Delta)\}\ra}

@@

\item	$\delta(e_1,e_2)$ represents a change, $\id{state}(e_2)\ominus\id{state}(e_1)$, for every pair $(e_1,e_2)\in\id{prec}$ (equation~\ref{def:delta});

@@

\section{theorems}
\label{sct:theorems}
	An order theoretic meet-semilattice $\la S, \leq\ra$ gives rise to a binary operation $\oplus$ such that $\la S, \oplus\ra$ is an algebraic meet-semilattice.
	With $\preceq$ being reflexive, antisymmetric, and transitive,
	it follows immediately that $\oplus$ is idempotent, commutative, and associative.
\section{Bibliography}
\bibliographystyle{elsarticle-harv}
\bibliography{doc}

>>>>

	We extend the notion of subset to historic structures and use the subset symbol $\subseteq$ for that purpose.
	Let us represent the fact that $h$ is a {\em historic substructure} of $h'$ by:
\begin{equation}
\label{def:historic substructure}
	\la \Events, \id{prec}\ra\ \subseteq\ \la \Events', \id{prec}'\ra\ \Leftrightarrow\ \Events\subseteq\Events'\ \wedge\ \id{prec}\subseteq\id{prec}'
\end{equation}

	A DIS grows by adding events to it.
	We describe each time a new event $e$ happens as a mapping, $\id{happens}_{P,e}$, from one history structure to another:
\begin{equation}
\begin{array}{l}
	P\subseteq\Events\ \wedge\ e\notin E\\
	\ \ \Rightarrow\ \ \id{happens}_{P,e}(\la \Events, \id{prec}\ra)\ =\ \la \Events\cup\{e\}, \id{prec}\cup\{(e_1,e) | e_1\in P\}\ra
\end{array}
\label{req:def happens}
\end{equation}
	This happening adds one event to $\Events$ and enlarges relation $\id{prec}$ with one pair for every immediate predecessor of $e$.
	The immediate predecessors of $e$ are specified in set $P$, which we call the predecessor set.
	We need the predecessor set to cater for the distributed nature of a DIS.
\begin{lemma}
\label{lemma:happens preserves history structure}
	If $\la \Events, \id{prec}\ra$ is a history structure and $P\subseteq\Events$ and $e\notin\Events$
	then $\id{happens}_{P,e}(\la \Events, \id{prec}\ra)$ is a history structure.
\end{lemma}
	Lemma~\ref{lemma:happens preserves history structure} is a preservation property.
	It says that $\id{happens}_{P,e}$ preserves the history structure,
	provided that $P\subseteq\Events$ and $e\notin\Events$.
	This lemma has been proven in proof assistant Isabelle/HOL.
	To understand why $P\subseteq\Events$ is necessary,
	let $\la \Events', \id{prec}'\ra=\id{happens}_{P,e}(h)$.
	Now suppose event $e_1\in P$ but $e_1\notin\Events$.
	Equation~\ref{req:def happens} tells us that $(e_1,e)\in\id{prec}'$.
	However, this violates requirement~\ref{req:precInE}.
	So it is necessary that every $e_1\in P$ is an element of $\Events$.
	Similarly, $e\notin\Events$ is a necessary condition,
	because any $e\in\Events$ violates the antisymmetry of $\id{prec}^+$ (requirement~\ref{req:precPlus asy}).

	With lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{happens}_{P,e}(h)=h'$ implies that $h\subseteq h'$.

@@

\section{States}
	This section defines the requirements for predictable behavior of a distributed information system (DIS).
	We consider behavior to be predictable if every state can be computed unambiguously from its history.
	For that purpose we introduce the notion of state.
	Let $\mathbb S$ represent the set of all conceivable states.

	This section expands history structures to stateful history substructures
	to represent the (global) state of an information system at any point in time.
	Since this paper is about distributed systems,
	we may not assume that the global state of a DIS at some point in time is observable.
	The state of a DIS is observable only in its local parts.
	So we shall define multiple states, each of which may be interpreted locally.

	We define a {\em stateful history structure} as a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
	$\la \Events, \id{prec}\ra$ is a history structure and $\id{state}:\Events\rightarrow{\mathbb S}$ is a function
	that maps every event in $\Events$ to a state.
	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful history structure.
	Let $s=\id{state}(e)$ represent the fact that $s$ is the state immediately after event $e$ has happened.
	Or shorter: $s$ is the state after event $e$.
%	Instead of $s=\id{state}(e)$, we may also write $(s,e)\in\id{state}$ because a function is a univalent and total relation
%	(and therefore a set of pairs).
	A {\em current state} is the state after a current event
	and a {\em past state} is the state after a past event.

	In a stateful history structure, the occurrence of an event $e$ defines a new history structure.
	This is the way systems evolve over time.
	So we define the occurrence of an event as a mapping, $\id{occurs}_{P,e,s}$, from one SHS to another SHS:
\begin{equation}
\begin{array}{l}
	P\subseteq\Events\ \wedge\ e\notin E\\
	\ \ \Rightarrow\ \ 
\begin{array}[t]{l}
	\id{occurs}_{P,e,s}(\la\Events, \id{prec}, \id{state}\ra)\\
	\ \ =\ \begin{array}[t]{@{}l@{\hspace{.2em}}c@{\hspace{.2em}}ll}
		\text{Let }\id{newState}(x)&=&s\text{,}&\text{if }x=e\\
				&=&\id{state}(x)\text{,}&\text{otherwise}\\
		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{newState}\}\ra}
		\end{array}
\end{array}
\end{array}
\label{req:def occurs}
\end{equation}
	This definition extends the definition of $\id{happens}$ (eqn.~\ref{req:def happens}).
	By using lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
	the stateful history structure.
	Also, we will use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em stateful historic substructure} of $h'$.
	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.
	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful history structure.
	To describe the growth of a stateful history structure by the event,
	we define $\id{SHS}$ recursively using $\id{occurs}_{P,e,s}$ as recurring step:
\begin{eqnarray}
	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
	\ \ \Rightarrow\ \id{SHS}(\id{occurs}_{P,e,s}\la \Events, \id{prec}, \id{state}\ra)
\end{array}
\label{req:recursive def occurs}
\end{eqnarray}

\section{Transactions}
\label{sct:Transactions}
	In practice, a new state is computed from the current state by specifying a change, $\Delta$.
	Such changes are called {\em transactions}.
	To describe a DIS in terms of changes, we need operators to compute a new state ($s_2$) from a given state ($s_1$).
	In fact, we need two operators, $\oplus$ and $\ominus$, such that
\begin{equation}
	s_2\ =\ s_1\oplus(s_2\ominus s_1)
\label{req:oplus and ominus}
\end{equation}
	Let us call $s_2\ominus s_1$ the difference ($\Delta$) between states $s_2$ and $s_1$.
	We use $s_1\oplus\Delta$ to compute a new state from an existing state $s_1$ and a change $\Delta$.

@@

	\ \ =\ \begin{array}[t]{@{}l@{\hspace{.2em}}c@{\hspace{.2em}}ll}

@@

		\text{Let }\id{newState}(x)&=&s_1\text{,}&\text{if }x=e\\
		&=&\id{state}(x)\text{,}&\text{otherwise}\\
		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{newState}\}\ra}

@@

\item	$\delta(e_1,e_2)$ represents a change, $\id{state}(e_2)\ominus\id{state}(e_1)$, for every pair $(e_1,e_2)\in\id{prec}$;

@@

\section{Durability}
\label{sct:Durability}
	In the eyes of the user, durability means that history is not changed.
	In the previous definitions durability has been secured by defining every change as an expansion of an existing system.
	Events are added only. Once added they are never removed or replaced.
	This corresponds to working practices in financial ledgers.
	Transactions are never removed, only compensated.
